Drug sensitivity prediction with normal inverse Gaussian shrinkage informed by external data

### 2020-04-06 docs/manuscript1.Rnw
fixed small mistake

<<analysis_gdsc>>==
# fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
# temp <- fit[, 1]
# fit <- as.matrix(fit[, -1])
# rownames(fit) <- temp
# 
# res <- read.table("results/analysis_gdsc_res1.txt", row.names=NULL)
# temp <- res[, 1]
# res <- as.matrix(res[, -1])
# rownames(res) <- temp
# 
# est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
# est1 <- round(est1, 2)
# est1[is.na(est1)] <- "-"
# colnames(est1) <- c("intercept", "not in pathway", "in pathway")
# est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)],
#               c(NA, 0, NA, 0, NA))
# est2 <- round(est2, 2)
# est2[is.na(est2)] <- "-"
# colnames(est2) <- c("intercept", "experimental", "development", "approved",
#                     "unkown", "targeted", "cytotoxic")
# 
# rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
#   c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
#     "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
# 
# tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
# tabm <- tab
# tabm <- round(tabm, 3)
# tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <-
#   paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
# colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
# rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
#   c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
#     "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")
# 
# library(glmnet)
# kurtosis <- function(x) {
#   n <- length(x)
#   kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) -
#     3*(n - 1)^2/((n - 2)*(n - 3))
#   replace(kurt, !is.finite(kurt), NA)
# }
# load("results/analysis_gdsc_fit1.Rdata")
# best2.semnig <- fit2.semnig$vb$mu
# best1.lasso <- lapply(fit1.lasso, function(s) {
#   unname(coef(s, "lambda.min")[-1, ])})
# best1.ridge <- lapply(fit1.ridge, function(s) {
#   unname(coef(s, "lambda.min")[-1, ])})
# best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
# kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE),
#            mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
#            mean(sapply(best1.ridge, kurtosis), na.rm=TRUE),
#            mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
# ranges <- rbind(range(unlist(lapply(1:length(
#   fit2.semnig$eb$mpriord), function(d) {
#     fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
#   fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
#     (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
ranges <- matrix(1, nrow=10, ncol=10)
tab <- matrix(1, nrow=10, ncol=10)
tabm <- matrix(1, nrow=10, ncol=10)
kurts <- matrix(1, nrow=10, ncol=10)
@
  The non-zero $\bm{\alpha}$ estimates in Tables 
  \ref{tab:analysis_gdsc_est1_feat} and \ref{tab:analysis_gdsc_est1_drug} show 
  that there is an effect of the
  external covariates. Models NIG$_{\text{f}+\text{d}}$, NIG$_{\text{f}}$, and
  bSEM estimate a negative effect for
  the external pathway covariate. This translates to a negative additive effect 
  on the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are 
  found
  in the target pathway of a drug are more predictive of IC50 than genes that
  are not in the target pathway, according to expectation. Furthermore the 
  NIG$_{\text{f}+\text{d}}$
  model estimates positive effects for the experimental and developmental drugs,
  compared to the approved drugs. This translates to smaller expected 
  prior precisions for the regression parameters of approved drugs, i.e.,
  approved drugs are easiest to predict, followed by developmental and 
  experimental drugs, as expected. On the other hand, the NIG$_{\text{d}}$ 
  model, that only includes external
  drug covariates, estimates a negative effect for the developmental drugs.
  The NIG$_{\text{f}+\text{d}}$ model estimates the smallest prior 
  $\bm{\beta}_d$ precision for the
  targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
  are assigned the largest prior $\bm{\beta}_d$ precision, according
  to expectation. Somewhat surprisingly, the NIG$_{\text{d}}$ model
  estimates the smallest
  prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
  targeted and unkown drugs. We conjecture that, due to absence of a local
  feature-specific variance component, this model is not flexible 
  enough to correctly capture the full covariance structure of the 
  $\bm{\beta}_d$ and therefore incorreclty estimates $\bm{\alpha}$.
  To see the total effect of the external covariates on the $\bm{\beta}_d$ 
  prior, we calculated the range of the expected prior variances 
  (up to a multiplicative error variance) as 
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[1, ]}\]$ and
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
  NIG$_{\text{f}+\text{d}}$ and bSEM, respectively. This confirms that the
  estimated prior 
  $\bm{\beta}_d$ variance components are in the same order of magnitude and that
  the external covariates have a detectable effect on the prior.
<<analysis_gdsc_est1_feat>>=
options(knitr.kable.NA="", digits=3)
kable(est1,
      col.names=colnames(est1), align="r",
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

>>>>

<<analysis_gdsc>>==
fit <- read.table("results/analysis_gdsc_fit1.txt", row.names=NULL)
temp <- fit[, 1]
fit <- as.matrix(fit[, -1])
rownames(fit) <- temp

res <- read.table("results/analysis_gdsc_res2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp

est1 <- cbind(fit[, 1], c(NA, 0, 0, NA, 0), fit[, 2])
est1 <- round(est1, 2)
est1[is.na(est1)] <- "-"
colnames(est1) <- c("intercept", "not in pathway", "in pathway")
est2 <- cbind(fit[, 3:5], c(NA, 0, NA, 0, NA), fit[, c(7:6)],
              c(NA, 0, NA, 0, NA))
est2 <- round(est2, 2)
est2[is.na(est2)] <- "-"
colnames(est2) <- c("intercept", "experimental", "development", "approved",
                    "unkown", "targeted", "cytotoxic")

rownames(est1)[c(1:4)] <- rownames(est2)[c(1:4)] <-
  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")

tab <- cbind(apply(res[rownames(res)=="pmse", ], 2, median))
tabm <- tab
tabm <- round(tabm, 3)
tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1] <-
  paste0("\\textbf{", tabm[which(tabm[, 1]==min(tabm[, 1], na.rm=TRUE)), 1], "}")
colnames(tabm) <- c("$\\text{PMSE}_{\\text{test}}$")
rownames(tabm)[c(1:4)] <- rownames(est2)[c(1:4)] <-
  c("NIG$_{\\text{f}+\\text{d}}^-$", "NIG$_{\\text{f}+\\text{d}}$",
    "NIG$_{\\text{f}}$", "NIG$_{\\text{d}}$")

library(glmnet)
kurtosis <- function(x) {
  n <- length(x)
  kurt <- (n + 1)*n*sum((x - mean(x))^4)/((n - 1)*(n - 2)*(n - 3)*var(x)^2) -
    3*(n - 1)^2/((n - 2)*(n - 3))
  replace(kurt, !is.finite(kurt), NA)
}
load("results/analysis_gdsc_fit1.Rdata")
best2.semnig <- fit2.semnig$vb$mu
best1.lasso <- lapply(fit1.lasso, function(s) {
  unname(coef(s, "lambda.min")[-1, ])})
best1.ridge <- lapply(fit1.ridge, function(s) {
  unname(coef(s, "lambda.min")[-1, ])})
best1.bSEM <- lapply(fit1.bSEM$vb$beta, function(s) {s[, 2]})
kurts <- c(mean(sapply(best2.semnig, kurtosis), na.rm=TRUE),
           mean(sapply(best1.lasso, kurtosis), na.rm=TRUE),
           mean(sapply(best1.ridge, kurtosis), na.rm=TRUE),
           mean(sapply(best1.bSEM, kurtosis), na.rm=TRUE))
ranges <- rbind(range(unlist(lapply(1:length(
  fit2.semnig$eb$mpriord), function(d) {
    fit2.semnig$eb$mpriord[d]*fit2.semnig$eb$mpriorf[[d]]}))),
  fit1.bSEM$eb$b[nrow(fit1.bSEM$eb$b), ]/
    (fit1.bSEM$eb$a[nrow(fit1.bSEM$eb$a), ] - 1))
ranges <- matrix(1, nrow=10, ncol=10)
tab <- matrix(1, nrow=10, ncol=10)
tabm <- matrix(1, nrow=10, ncol=10)
kurts <- matrix(1, nrow=10, ncol=10)
@
  The non-zero $\bm{\alpha}$ estimates in Tables 
  \ref{tab:analysis_gdsc_est1_feat} and \ref{tab:analysis_gdsc_est1_drug} show 
  that there is an effect of the
  external covariates. Models NIG$_{\text{f}+\text{d}}$, NIG$_{\text{f}}$, and
  bSEM estimate a negative effect for
  the external pathway covariate. This translates to a negative additive effect 
  on the expected prior precision of the $\bm{\beta}_d$, i.e., genes that are 
  found
  in the target pathway of a drug are more predictive of IC50 than genes that
  are not in the target pathway, according to expectation. Furthermore the 
  NIG$_{\text{f}+\text{d}}$
  model estimates positive effects for the experimental and developmental drugs,
  compared to the approved drugs. This translates to smaller expected 
  prior precisions for the regression parameters of approved drugs, i.e.,
  approved drugs are easiest to predict, followed by developmental and 
  experimental drugs, as expected. On the other hand, the NIG$_{\text{d}}$ 
  model, that only includes external
  drug covariates, estimates a negative effect for the developmental drugs.
  The NIG$_{\text{f}+\text{d}}$ model estimates the smallest prior 
  $\bm{\beta}_d$ precision for the
  targeted drugs, followed by cytotoxic drugs, while drugs with an unkown target
  are assigned the largest prior $\bm{\beta}_d$ precision, according
  to expectation. Somewhat surprisingly, the NIG$_{\text{d}}$ model
  estimates the smallest
  prior $\bm{\beta}_d$ precision for the cytotoxic drugs, compared to the 
  targeted and unkown drugs. We conjecture that, due to absence of a local
  feature-specific variance component, this model is not flexible 
  enough to correctly capture the full covariance structure of the 
  $\bm{\beta}_d$ and therefore incorreclty estimates $\bm{\alpha}$.
  To see the total effect of the external covariates on the $\bm{\beta}_d$ 
  prior, we calculated the range of the expected prior variances 
  (up to a multiplicative error variance) as 
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[1, ]}\]$ and
  $\hat{E}(\gamma_{jd}^2 \cdot \tau_d^2) \in \[ \Sexpr{ranges[2, ]}\]$ for
  NIG$_{\text{f}+\text{d}}$ and bSEM, respectively. This confirms that the
  estimated prior 
  $\bm{\beta}_d$ variance components are in the same order of magnitude and that
  the external covariates have a detectable effect on the prior.
<<analysis_gdsc_est1_feat>>=
options(knitr.kable.NA="", digits=3)
kable(est1,
      col.names=colnames(est1), align="r",
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@
